TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2018-08-13 paper/draft.tex
accuracy in the data space

\begin{abstract}
\thecannon\ is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution, near-infrared \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
The lack of reliable stellar models in the near infrared for temperatures $<3500$\,K motivates the extension of this approach to M-dwarf stars, but with new training labels from optical surveys that overlap \sdssiv\ \apogee.
Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, and demonstrate that \thecannon\ acting on \apogee\ spectra can infer spectral types with predictive accuracy of 0.9 types.
The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee, selected in part on \gaia\ photometry and astrometry.
We validate our results using photometric temperature indicators and the \gaia\ color--magnitude diagram.
As a side result, we determine empirical relationships between spectral type and temperature and metallicity.
These results have implications for the NASA \tess\ Mission and other projects that involve large samples of M dwarfs.
\end{abstract}

@@

In this work we build a data-driven model for M dwarf spectra with \thecannon (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}).
\thecannon is a regression (typically quadratic) for the spectral pixels given a training
set of stars with reliable or accurate or assumed-accurate labels.
The model predicts spectrum given labels, but it can be turned around and also
used to predict labels given spectrum.
Thus it is a generative model for stellar spectra that can be used for parameter inference.

@@

\color{gcolor}{HOGG: Some commentary on why $\chi^2$ is higher than expected.}\color{black}

>>>>

\begin{abstract}
\thecannon\ is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution, near-infrared \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
The lack of reliable stellar models in the near infrared for temperatures $<3500$\,K motivates the extension of this approach to M-dwarf stars, but with new training labels from optical surveys that overlap \sdssiv\ \apogee.
Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
These are parameter-space precision statements, but the model is accurate in the data space: It accurately predicts \apogee\ spectral pixels at the percent level.
We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, and demonstrate that \thecannon\ acting on \apogee\ spectra can infer spectral types with predictive accuracy of 0.9 types.
The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee, selected in part on \gaia\ photometry and astrometry.
We validate our results using photometric temperature indicators and the \gaia\ color--magnitude diagram.
As a side result, we determine empirical relationships between spectral type and temperature and metallicity.
These results have implications for the NASA \tess\ Mission and other projects that involve large samples of M dwarfs.
\end{abstract}

@@

In this work we build a data-driven model for M dwarf spectra with \thecannon\ (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}).
\thecannon\ is a regression (typically quadratic) for the spectral pixels given a training
set of stars with reliable or accurate or assumed-accurate labels.
The model predicts spectrum given labels, but it can be turned around and also
used to predict labels given spectrum.
Thus it is a generative model for stellar spectra that can be used for parameter inference.

@@

\color{gcolor}{HOGG: Some commentary on why $\chi^2$ is higher than expected. Also explain why this means that the model is good to percent level.}\color{black}
